Some examples in extended logic counting classes

### 2017-01-12 LICS/appendix-syntactic.tex
Version nueva de minus one.

\noindent {\em Part (1).}  Let $\alpha =  \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\, \varphi(\bar{X},\bar{x},\bar{y})$ where $\varphi$ is quantifier-free and with  atomic formulas in $\fo$. We call this syntax {\em extended quantifier-free} from now on. 
Note that, if $\alpha$ is of the form $\alpha = \sa{\bar{x}}\exists\bar{y}\,\varphi(\bar{x},\bar{y})$ (i.e. $\length(\bar{X}) = 0$), we can define $\kappa(\alpha) = \sa{\bar{x}}\exists\bar{y}[\varphi(\bar{x},\bar{y})\wedge\exists\bar{z}(\varphi(\bar{z},\bar{y})\wedge\bar{z} < \bar{x})]$, which is in $\eqso(\logex{1})$ and fulfills the desired condition. 
Therefore, for the rest of the proof we can assume that $\length(\bar{X}) > 0$ and $\length(\bar{x}) > 0$.

 Let $\bar{x} = (x_1,\ldots,x_{\length(\bar{x})})$ and $\bar{X} = (X_1,\ldots,X_{\length(\bar{X})})$.
To simplify the analysis of $\varphi$, the first step is to rewrite $\varphi$ in into a DNF formula. 
More precisely, we rewrite $\varphi$ into an equivalent formula of the form $\bigvee_{i = 1}^m \varphi_i$ for some $m\in \bbN$ where each $\varphi_i$ is a conjunction of atomic formulas or negation of atomic formulas. 
Furthermore, we suppose that each conjunct $\varphi_i$ has the~form:
$$
\varphi_i(\bar{X},\bar{x},\bar{y}) = \underbrace{\varphi_i^{\fo}(\bar{x},\bar{y})}_{\text{an $\fo$ formula}} \wedge 
\underbrace{\varphi_i^{+}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $X_j$'s}} \wedge
\underbrace{\varphi_i^{-}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $\neg X_j$'s}}.
$$
Note that atomic formulas of the form $R(\bar{z})$ where $R \in \R$ can appear in the subformula $\varphi_i^{\fo}(\bar{x},\bar{y})$.

Now, we define a series of rewritings to $\varphi$ that will make each formula $\varphi_i$ satisfy the following three conditions: (a) each first-order variable in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$ is mentioned at most once and, moreover, no variable in $\bar{x}$ is mentioned, (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines an ordered partition over the variables in $(\bar{x},\bar{y})$ (see below for the definition of ordered partition) and (c) if $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ are mentioned, then the ordered partition should not satisfy $\bar{z} = \bar{w}$.
We explain below how to rewrite $\varphi_i$ in order to satisfy each condition.
\begin{itemize}
	\item[(a)] In order to satisfy the first condition, consider some instance of a $X_j(\bar{w})$ in $\varphi_i$, where $\bar{w}$ is a subtuple of $(\bar{x},\bar{y})$. We add $\length(\bar{w})$ new variables $z_1,\ldots,z_{\length(\bar{w})}$ to the formula and let $\bar{z} = (z_1,\ldots,z_{\length(\bar{w})})$. We redefine $\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$ by replacing $X_j(\bar{w})$ with $X_j(\bar{z})$. The formula $\varphi_i$ is now defined as
	$$
	\varphi_i(\bar{X},\bar{x},(\bar{y},\bar{z})) = (\bar{z} = \bar{w} \wedge \varphi_i^{\fo}(\bar{x},\bar{y})) \wedge \varphi_i^{-}(\bar{X},\bar{x},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{x},\bar{y}).
	$$
	We repeat this process for each instance of a $X_j(\bar{w})$ in $\varphi_i$, and we obtain a formula where none of the $X_j$'s acts over any variable in $\bar{x}$. We add the new first-order variables to $\bar{y}$ and we redefine $\varphi_i$ as
	$$
	\varphi_i(\bar{X},\bar{x},\bar{y}) = \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y}).
	$$
	For example, if $\bar{x} = x$, $\bar{y} = y$ and $\varphi_i = X(x,y)\wedge \neg X(x,x) \wedge x < y$, then we redefine $\bar{y} = (y,v_1,v_2,v_3,v_4)$ and $\varphi_i := X(v_1,v_2) \wedge \neg X(v_3,v_4) \wedge x < y \wedge v_1 = x \wedge v_2 = y \wedge v_3 = x \wedge v_4 = x.$ 
	\item[(b)] An ordered partition on a set $S$ is defined by an equivalence relation $\sim$ over $S$, and a linear order over $S/\!\sim$. For example, let $\bar{x} = (x_1,x_2,x_3,x_4)$. A possible ordered partition would be defined by the formula $\theta(\bar{x}) = x_2 < x_1 \wedge x_1 = x_4 \wedge x_4 < x_3$. On the other hand, the formula $\theta'(\bar{x}) = x_1 < x_2 \wedge x_1 < x_4 \wedge x_2 = x_3$ does not define an ordered partition since both $\{x_1\}<\{x_2,x_3\}<\{x_4\}$ and $\{x_1\} < \{x_2,x_3,x_4\}$ satisfy $\theta'$.
	For a given $k$, let $\cB_k$ be the number of possible ordered partitions for a set of size $k$. For $1 \leq j \leq \cB_{\length(\bar{x},\bar{y})}$ 
	let $\theta^j(\bar{x},\bar{y})$ be a formula that defines an ordered partition over $(\bar{x},\bar{y})$.

	Thus, the formula $\varphi(\bar{X},\bar{x},\bar{y})$ is then redefined as:
	$$
	\bigvee_{i = 1}^m \bigvee_{j = 1}^{\cB_{\length(\bar{x},\bar{y})}}[\theta^j(\bar{x},\bar{y})\wedge \varphi_i(\bar{X},\bar{x},\bar{y})],
	$$
	Note that each $\theta^j(\bar{x},\bar{y})$ is an $\fo$-formula.
	Then, by redefining the conjuncts $\varphi_i(\bar{X},\bar{x},\bar{y})$ of $\varphi$ to $\theta^j(\bar{x},\bar{y})\wedge \varphi_i(\bar{X},\bar{x},\bar{y})$, we can suppose that each $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines an ordered partition over the variables in $(\bar{x},\bar{y})$.
	\item[(c)] For showing that no $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ are mentioned in $\varphi_i$ with $\bar{z}$ and $\bar{w}$ equivalent in the ordered partition (i.e. $\varphi_i$ is inconsistent), we do the following. If there exists an instance of $X_j(\bar{z})$ in $\varphi^{+}_i$, an instance of $\neg X_j(\bar{w})$ in $\varphi^{-}_i$ and the ordered partition in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$, then the entire formula $\varphi_i$ is removed from $\varphi$.
\end{itemize}
It is important to notice that the resulting $\varphi$ is equivalent to the initial one, and it is still an extended quantifier-free formula. From now on, we assume that each $\varphi_i(\bar{X},\bar{x},\bar{y})$ satisfies conditions (a), (b) and (c). Furthermore, by condition (a) we have that:
$$
\varphi_i(\bar{X},\bar{x},\bar{y}) \; = \; \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y}) 
$$
where $\varphi^{+}$ and $\varphi^{-}_i$ do not depend on the variables $\bar{x}$.
\begin{claim}\label{claim:minusone}
For a given ordered structure $\A$, if there exists a first-order assignment $v$ for $\A$ such that $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$, then there exists a second-order assignment $V$ for $\A$ such that  $(\A,V,v)\models\varphi_i(\bar{X},\bar{x},\bar{y})$.
\end{claim}
\begin{proof}
Let $\A$ be an ordered structure with domain $A$ and let $v$ be a first-order assignment for $\A$ such that $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$.
Define $\bar{B} = (B_1,\ldots,B_{\length(\bar{X})})$ as $B_j = \{v(\bar{w})\mid \text{ $X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$, and let $V$ be a second-order assignment for which $V(\bar{X}) = \bar{B}$.
Towards a contradiction, suppose that $(\A,V,v)\not\models\varphi_i(\bar{X},\bar{x},\bar{y})$.
By the choice of $v$, and construction of $V$ it is clear that $(\A,V,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})\wedge\varphi^{+}_i(\bar{X},\bar{x},\bar{y})$, so we necessarily have that $(\A,V,v)\not\models\varphi^{-}_i(\bar{X},\bar{x},\bar{y})$.
Let $X_t$ be such that $\neg X_t(\bar{w})$ is mentioned in $\varphi^{-}_i(\bar{X},\bar{y})$ and $(\A,V, v)\not\models\neg X_t(\bar{w})$, namely, $v(\bar{w})\in B_t$. 
However, by the construction of $B_t$, there exists a subtuple $\bar{z}$ of $\bar{y}$ such that $X_t(\bar{z})$ appears in $\varphi^{+}_i(\bar{X},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$. Since $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$, then the ordered partition in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$. This violates condition (c) since $\neg X_t(\bar{w})$ appears in $\varphi^{-}_i$ and $X_t(\bar{z})$ appears in $\varphi^{+}_i$, which leads to a contradiction. 
\end{proof}

The previous claim and proof motivates the following definitions, in particular, a generalization of the sequence $\bar{B}$. 
For a structure $\A$ and a first-order assignment $v$ for $\A$, define $\bar{B}^v = (B^v_1,\ldots,B^v_{\length(\bar{X})})$ where each $B^v_j = \{v(\bar{w}) \mid \text{$X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$. 
Moreover, let $\bar{v}$ be the sequence of all different first-order variable mentioned in $\varphi_i^{+}$ and let $\bar{u}$ be a sequence of variables disjoint from $\bar{v}$ such that $(\bar{v},\bar{u})$ and $(\bar{x},\bar{y})$ have the same set of variables. Then define $v^*$ to be the first-order assignment for $\A$ for which we have that $v^{*}(\bar{v})$ is the lexicographically smallest tuple that satisfies $(\A,v^{*})\models\exists\bar{u} \varphi^{\fo}_i(\bar{u},\bar{v})$.
Indeed, one ca define a formula $\mu_i$ that defines $v^*$ as follows:
$$
\mu_i(\bar{v}) = \exists\bar{u}\,\varphi_i^{\fo}(\bar{u},\bar{v})\wedge\forall\bar{u}'\forall\bar{v}'(\varphi_i^{\fo}(\bar{u}',\bar{v}')\to\bar{v}\leq\bar{v}').
$$ 
In other words, we have that $(\A,v)\models \mu_i(\bar{v})$ if, and only if, $v(\bar{v}) = v^{*}(\bar{v})$ for every assignment $v$. 
Finally, define $\bar{B}^* = \bar{B}^{v^{*}}$ and $V^*$ as some second-order assignment for $\A$ for which $V^*(\bar{X}) = \bar{B}^*$.

Now, we have all the tools to introduce the main formula of the proof. 
Intuitively, the pair $(V^*, v^*)$ is the minimum assignment that satisfies $\varphi_i(\bar{X}, \bar{x})$ and our candidate to be removed from the semantics of $\varphi(\bar{X}, \bar{x})$. Towards this goal, we can define a formula $\psi_i(\bar{X},\bar{x})$ that has the following key property: 
\begin{center}
if $\varphi_i(\bar{X}, \bar{x}, \bar{y})$ is satisfiable, then $(\A,V,v) \models\varphi_i(\bar{X},\bar{x},\bar{y})$ if, and only if, $V(\bar{X}) \neq V^*(\bar{X})$ or $v(\bar{x}) \neq v^*(\bar{x})$ for every $(V,v)$. 
\end{center}
That is, its exclude somehow the pair $(V^*, v^*)$. This property can be defined in logic as follows:
\begin{align}
\psi_i(\bar{X},\bar{x}) = & \;\; \big( \, \exists\bar{x}\exists\bar{y}\, \varphi^{\fo}_i(\bar{x},\bar{y}) \,\big) \rightarrow  \label{eq:part1} \\
& \hspace{1cm} \Big( \; 
\exists\bar{x}'(\exists\bar{y}'\varphi_i(\bar{X},\bar{x}',\bar{y}')\wedge \bar{x}'<\bar{x}) \;\;  \vee   \label{eq:part2} \\
& \hspace{1.5cm} \exists\bar{v} (\mu_i(\bar{v})\wedge \big( \varphi^{+}_i(\bar{X},\bar{v}) \rightarrow \bigvee_{X \in \bar{X}} \exists\bar{z}\, ( \, X(\bar{z}) \wedge \bigwedge\limits_{X(\bar{w}) \, \in \, \varphi^{+}_i(\bar{X},\bar{v})}\bar{w}\neq\bar{z} \, ) \big) \Big)  \label{eq:part3}
\end{align}
To understand the formula, first notice that the premise of the implication at (\ref{eq:part1}) is true if, and only if, $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. 
Indeed, by Claim~\ref{claim:minusone} we know that if $\exists\bar{x}\exists\bar{y}\, \varphi^{\fo}_i(\bar{x},\bar{y})$ is true, then there exists assignments $V$ and $v$ such that $(\A,V,v) \models\varphi_i(\bar{X},\bar{x},\bar{y})$.
Then, the conclusion of the implication at (\ref{eq:part2}) and (\ref{eq:part3}), take care that the $\varphi_i$ is only satisfied by a pair $(V, v)$ different from $(V^*, v^*)$.
Here, the first disjunct at (\ref{eq:part2}) check whether $(\A,V,v) \models\varphi_i(\bar{X},\bar{x},\bar{y})$ but $v(\bar{x})$ is not the lexicographically smallest tuple that have this property (i.e. $v(\bar{x}) \neq v^*(\bar{x})$).
The second disjunct at (\ref{eq:part3}) is more involved and it can be explained as follows:

The following is a pivotal formula in our proof:

\begin{align}
\psi_i(\bar{X},\bar{x}) = &\exists\bar{v}\bigg(\overbrace{\mu_i(\bar{v})}^{(\blacktriangleleft)}\wedge\Big( \overbrace{\neg\varphi^{+}_i(\bar{X},\bar{v})}^{(\blacktriangle)} \vee \overbrace{\bigvee_{X \in \bar{X}} \exists\bar{z}(X(\bar{z}) \wedge \bigwedge\limits_{\substack{\text{subtuples }\bar{w} \text{ of }\bar{v} \text{ s.t.}\\ X(\bar{w})\text{ is mentioned} \\ \text{in }\varphi^{+}_i(\bar{X},\bar{v})}}\bar{w}\neq\bar{z}}^{(\blacktriangledown)})\Big)\bigg) \vee \label{minusone-1}\\ &\exists\bar{x}'(\exists\bar{y}'\varphi_i(\bar{X},\bar{x}',\bar{y}')\wedge \bar{x}'<\bar{x})\vee \label{minusone-2} \\ 
&\neg\exists\bar{x}\exists\bar{y}\varphi^{\fo}_i(\bar{x},\bar{y}). \label{minusone-3}
\end{align}

Consider some $\A$. In detail, the formula \eqref{minusone-1} represents the following: In $(\blacktriangleleft)$, we make it so the only value for $\bar{v}$ we consider is $v^{*}(\bar{v})$. In $(\blacktriangle)$, we mention that the assignment for $\bar{X}$ satisfies the formula if does not contain at least one of the tuples in $v^{*}(\bar{v})$. In $(\blacktriangledown)$, we mention that in the assignment for $\bar{X}$ also satisfies the formula if it contains at least one tuple, in one of the $X$'s, that is not equal to the one in $v^{*}(\bar{v})$. Then, in the formula \eqref{minusone-2} we mention that the assignment for $\bar{x}$ satisfies the formula if it is not the lexicographically smallest assignment that satisfies $\varphi_i$ for that particular assignment for $\bar{X}$. Lastly, in \eqref{minusone-3}, we mention that the assignment for $(\bar{X},\bar{x})$ satisfies the formula if there are no assignments $V,v$ for $\A$ such that $(\A,V,v)\not\models \exists\bar{y}\varphi_i(\bar{X},\bar{x},\bar{y})$.

>>>>

\noindent {\em Part (1).}  Let $\alpha =  \sa{\bar{X}}\sa{\bar{x}} \varphi(\bar{X},\bar{x})$ where $\varphi$ is an  $\fo$-formula.
Note that, if $\alpha$ is of the form $\alpha = \sa{\bar{x}} \varphi(\bar{x})$ (i.e. $\length(\bar{X}) = 0$), we can define $\kappa(\alpha) = \sa{\bar{x}} [\varphi(\bar{x})\wedge\exists\bar{z}(\varphi(\bar{z})\wedge\bar{z} < \bar{x})]$, which is in $\eqso(\logex{1})$ and fulfills the desired condition. 
Therefore, for the rest of the proof we can assume that $\length(\bar{X}) > 0$ and $\length(\bar{x}) > 0$.

 Let $\bar{x} = (x_1,\ldots,x_{\length(\bar{x})})$ and $\bar{X} = (X_1,\ldots,X_{\length(\bar{X})})$.
To simplify the analysis of $\varphi$, the first step is to rewrite $\varphi$ in into a DNF formula. 
More precisely, we rewrite $\varphi$ into an equivalent formula of the form $\bigvee_{i = 1}^m \varphi_i$ for some $m\in \bbN$ where each $\varphi_i$ is a conjunction of atomic formulas or negation of atomic formulas. 
Furthermore, we suppose that each conjunct $\varphi_i$ has the~form:
$$
\varphi_i(\bar{X},\bar{x}) = \exists \bar{y} \; \underbrace{\varphi_i^{\fo}(\bar{x},\bar{y})}_{\text{an $\fo$ formula}} \wedge 
\underbrace{\varphi_i^{+}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $X_j$'s}} \wedge
\underbrace{\varphi_i^{-}(\bar{X},\bar{x},\bar{y})}_{\text{conjunction of $\neg X_j$'s}}.
$$
Note that atomic formulas of the form $R(\bar{z})$ where $R \in \R$ can appear in the subformula $\varphi_i^{\fo}(\bar{x},\bar{y})$.

Now, we define a series of rewritings to $\varphi$ that will make each formula $\varphi_i$ satisfy the following three conditions: (a) no variable from $\bar{x}$ are mentioned in $\varphi_i^{-}(\bar{X},\bar{x},\bar{y})\wedge\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$, (b) $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines an ordered partition over the variables in $(\bar{x},\bar{y})$ (see below for the definition of ordered partition) and (c) if $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ are mentioned, then the ordered partition should not satisfy $\bar{z} = \bar{w}$.
We explain below how to rewrite $\varphi_i$ in order to satisfy each condition.
\begin{itemize}
	\item[(a)] In order to satisfy the first condition, consider some instance of a $X_j(\bar{w})$ in $\varphi_i$, where $\bar{w}$ is a subtuple of $(\bar{x},\bar{y})$. We add $\length(\bar{w})$ new variables $z_1,\ldots,z_{\length(\bar{w})}$ to the formula and let $\bar{z} = (z_1,\ldots,z_{\length(\bar{w})})$. We redefine $\varphi_i^{+}(\bar{X},\bar{x},\bar{y})$ by replacing $X_j(\bar{w})$ with $X_j(\bar{z})$. The formula $\varphi_i$ is now defined as
	$$
	\varphi_i(\bar{X},\bar{x}) = \exists \bar{y} \exists \bar{z} \; (\bar{z} = \bar{w} \wedge \varphi_i^{\fo}(\bar{x},\bar{y})) \wedge \varphi_i^{-}(\bar{X},\bar{x},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{x},\bar{y})).
	$$
	We repeat this process for each instance of a $X_j(\bar{w})$ in $\varphi_i$, and we obtain a formula where none of the $X_j$'s acts over any variable in $\bar{x}$. We add the new first-order variables to $\bar{y}$ and we redefine $\varphi_i$ as
	$$
	\varphi_i(\bar{X},\bar{x}) = \exists \bar{y} \; \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y}).
	$$
	For example, if $\bar{x} = x$, $\bar{y} = y$ and $\varphi_i = \exists \bar{y} \;  X(x,y)\wedge \neg X(x,x) \wedge x < y$, then we redefine $\bar{y} = (y,v_1,v_2,v_3,v_4)$ and $\varphi_i := \exists \bar{y} \;  X(v_1,v_2) \wedge \neg X(v_3,v_4) \wedge x < y \wedge v_1 = x \wedge v_2 = y \wedge v_3 = x \wedge v_4 = x.$ 
	\item[(b)] An ordered partition on a set $S$ is defined by an equivalence relation $\sim$ over $S$, and a linear order over $S/\!\sim$. For example, let $\bar{x} = (x_1,x_2,x_3,x_4)$. A possible ordered partition would be defined by the formula $\theta(\bar{x}) = x_2 < x_1 \wedge x_1 = x_4 \wedge x_4 < x_3$. On the other hand, the formula $\theta'(\bar{x}) = x_1 < x_2 \wedge x_1 < x_4 \wedge x_2 = x_3$ does not define an ordered partition since both $\{x_1\}<\{x_2,x_3\}<\{x_4\}$ and $\{x_1\} < \{x_2,x_3,x_4\}$ satisfy $\theta'$.
	For a given $k$, let $\cB_k$ be the number of possible ordered partitions for a set of size $k$. For $1 \leq j \leq \cB_{\length(\bar{x},\bar{y})}$ 
	let $\theta^j(\bar{x},\bar{y})$ be a formula that defines an ordered partition over $(\bar{x},\bar{y})$.

	Thus, the formula $\varphi_i(\bar{X},\bar{x})$ is then redefined as:
	$$
	\bigvee_{i = 1}^m \bigvee_{j = 1}^{\cB_{\length(\bar{x},\bar{y})}} \exists \bar{y} \; [\theta^j(\bar{x},\bar{y})\wedge \varphi_i^{\fo}(\bar{x},\bar{y}) \wedge \varphi_i^{-}(\bar{X},\bar{y}) \wedge \varphi_i^{+}(\bar{X},\bar{y})],
	$$
	Note that each $\theta^j(\bar{x},\bar{y})$ is an $\fo$-formula.
	Then, by redefining $\varphi_i^{\fo}(\bar{x},\bar{y})$  to $\theta^j(\bar{x},\bar{y})\wedge \varphi_i^{\fo}(\bar{x},\bar{y})$, we can suppose that each $\varphi_i^{\fo}(\bar{x},\bar{y})$ defines an ordered partition over the variables in $(\bar{x},\bar{y})$.
	\item[(c)] For showing that no $X_j(\bar{z})$ and $\neg X_j(\bar{w})$ are mentioned in $\varphi_i$ with $\bar{z}$ and $\bar{w}$ equivalent in the ordered partition (i.e. $\varphi_i$ is inconsistent), we do the following. If there exists an instance of $X_j(\bar{z})$ in $\varphi^{+}_i$, an instance of $\neg X_j(\bar{w})$ in $\varphi^{-}_i$ and the ordered partition in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$, then the entire formula $\varphi_i$ is removed from $\varphi$.
\end{itemize}
It is important to notice that the resulting $\varphi$ is equivalent to the initial one, and it is still a formula in $\eqso(\logex{1})$. From now on, we assume that each $\varphi_i(\bar{X},\bar{x},\bar{y})$ satisfies conditions (a), (b) and (c). Furthermore, it has the following form:
$$
\varphi_i(\bar{X},\bar{x}) \; = \; \exists{\bar{y}} \; \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y}) 
$$
where $\varphi^{+}$ and $\varphi^{-}_i$ do not depend on  $\bar{x}$.
\begin{claim}\label{claim:minusone}
For a given ordered structure $\A$, if there exists a first-order assignment $v$ for $\A$ such that $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$, then there exists a second-order assignment $V$ for $\A$ such that  $(\A,V,v)\models \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$.
\end{claim}
\begin{proof}
Let $\A$ be an ordered structure with domain $A$ and let $v$ be a first-order assignment for $\A$ such that $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$.
Define $\bar{B} = (B_1,\ldots,B_{\length(\bar{X})})$ as $B_j = \{v(\bar{w})\mid \text{ $X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$, and let $V$ be a second-order assignment for which $V(\bar{X}) = \bar{B}$.
Towards a contradiction, suppose that $(\A,V,v)\not\models \varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$.
By the choice of $v$, and construction of $V$ it is clear that $(\A,V,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})\wedge\varphi^{+}_i(\bar{X},\bar{y})$, so we necessarily have that $(\A,V,v)\not\models\varphi^{-}_i(\bar{X},\bar{y})$.
Let $X_t$ be such that $\neg X_t(\bar{w})$ is mentioned in $\varphi^{-}_i(\bar{X},\bar{y})$ and $(\A,V, v)\not\models\neg X_t(\bar{w})$, namely, $v(\bar{w})\in B_t$. 
However, by the construction of $B_t$, there exists a subtuple $\bar{z}$ of $\bar{y}$ such that $X_t(\bar{z})$ appears in $\varphi^{+}_i(\bar{X},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$. Since $(\A,v)\models\varphi^{\fo}_i(\bar{x},\bar{y})$ and $v(\bar{z}) = v(\bar{w})$, then the ordered partition in $\varphi^{\fo}_i$ satisfies $\bar{z} = \bar{w}$. This violates condition (c) since $\neg X_t(\bar{w})$ appears in $\varphi^{-}_i$ and $X_t(\bar{z})$ appears in $\varphi^{+}_i$, which leads to a contradiction. 
\end{proof}

The previous claim and proof motivates the following definitions.
For a structure $\A$ and a first-order assignment $v$ for $\A$, define $\bar{B}^v = (B^v_1,\ldots,B^v_{\length(\bar{X})})$ where each $B^v_j = \{v(\bar{w}) \mid \text{$X_j(\bar{w})$ is mentioned in $\varphi^{+}_i(\bar{X},\bar{y})$}\}$.
One can easily check that for every assignments $V, v$ such that $(\A,V,v)\models\varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$, then it holds that $(\A,\bar{B}^v,v)\models\varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$ and $\bar{B}^v \subseteq V(\bar{X})$, namely, $\bar{B}^v$ is a valid candidate for $\bar{X}$ and, furthermore, it is contained in all the assignment of $\bar{X}$ when $v$ is fixed.
Therefore, by choosing one particular $v$ the plan is to remove $\bar{B}^v$ as an assignment over $\bar{X}$ in $\varphi_i$. 
For this, the idea is to choose the minimal $v$ that satisfies $\varphi^{\fo}_i(\bar{x},\bar{y})$ which can be defined with the following formula:
$$
\text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \; = \;  \,\varphi_i^{\fo}(\bar{x},\bar{y})\wedge\forall\bar{x}'\forall\bar{y}'(\varphi_i^{\fo}(\bar{x}',\bar{y}')\to ( \bar{x}\leq\bar{x}' \wedge  \bar{y}\leq\bar{y}')).
$$ 
If $\varphi^{\fo}_i$ is satisfiable, let $v$ be the only assignment such that  $(\A,v)\models \text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y})$. 
Furthermore, let $V^*$ the second order assignment that satisfies $V^*(\bar{X}) = \bar{B}^{v}$ and $v^*$ the first order assignment such that $v^*(\bar{x}) = v(\bar{x})$.
By the previous discussion, we have that $(\A,V^*,v^*)\models\varphi_i(\bar{X},\bar{x})$.

Now, we have all the tools to introduce the main formula of the proof. 
Intuitively, we want to exclude the assignment $(V^*, v^*)$ as a satisfying assignment for $\varphi_i(\bar{X},\bar{x})$.
Towards this goal, we can define a formula $\psi_i(\bar{X},\bar{x})$ such that $(\A, V, v) \models \psi_i(\bar{X},\bar{x})$ if, and only if, 
if $\varphi_i(\bar{X}, \bar{x})$ is satisfiable, then $V \neq V^*$ or $v \neq v^*$. 
That is, it excludes all pairs $(V, v)$ that coincide with $(V^*, v^*)$ in $(\bar{X}, \bar{x})$ whenever $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. This property can be defined as follows:
\begin{align}
\psi_i(\bar{X},\bar{x}) = & \;\; \big( \, \exists\bar{x}\exists\bar{y}\, \varphi^{\fo}_i(\bar{x},\bar{y}) \,\big) \rightarrow  \label{eq:part1} \\
& \hspace{1cm} \Big( \; 
\exists\bar{y} \;\; \text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \wedge \big( \varphi_i^+(\bar{X},\bar{x}, \bar{y})  \rightarrow \bigvee_{X \in \bar{X}} \exists\bar{z}\, ( \, X(\bar{z}) \wedge \bigwedge\limits_{X(\bar{w}) \, \in \, \varphi^{+}_i(\bar{X},\bar{v})}\bar{w}\neq\bar{z} \, ) \big)  \;\;\;\;  \vee   \label{eq:part2} \\ 
& \hspace{1.5cm} (\exists\bar{x}'\exists\bar{y}\, \varphi_i(\bar{X},\bar{x}',\bar{y})\wedge \bar{x}' < \bar{x}) \;\; \Big)  \label{eq:part3}
\end{align}
To understand the formula, first notice that the premise of the implication at (\ref{eq:part1}) is true if, and only if, $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. 
Indeed, by Claim~\ref{claim:minusone} we know that if $\exists\bar{x}\exists\bar{y}\, \varphi^{\fo}_i(\bar{x},\bar{y})$ is true, then there exists assignments $V$ and $v$ such that $(\A,V,v) \models\varphi^{\fo}_i(\bar{x},\bar{y}) \wedge \varphi^{+}_i(\bar{X}, \bar{y}) \wedge \varphi^{-}_i(\bar{X},\bar{y})$.
Then, the conclusion of the implication at (\ref{eq:part2}) and (\ref{eq:part3}), take care that $V(\bar{X}) \neq V^*(\bar{X})$ or $v(\bar{x}) \neq v^*(\bar{x})$.
Here, the first disjunct (\ref{eq:part2}) checks that $V(\bar{X}) \neq V^*(\bar{X})$ by showing that if $\varphi_i(\bar{X},\bar{x})$ is satisfied then $V(\bar{X})$ must contain one element more that it is not in $V^*(\bar{X})$. 
The second disjunct (\ref{eq:part3}) is satisfied when $v(\bar{x})$ is not the lexicographically smallest tuple that satisfies $\varphi_i$ (i.e. $v(\bar{x}) \neq v^*(\bar{x})$).

We are ready to define the formula  $\kappa(\alpha) = \sa{\bar{X}}\sa{\bar{x}} \bigvee_{i = 1}^m\varphi_i'(\bar{X},\bar{x})$
where each modified disjunct $\varphi_i'(\bar{X},\bar{x})$ is constructed as follows. 
For the sake of simplification, define the auxiliary formula $\chi_i = \neg\exists\bar{x}\exists\bar{y}\,\varphi^{\fo}_i(\bar{x},\bar{y})$. 
This formula basically checks if $\varphi_i$ is satisfiable or not.
Then, first define the formula:
$$
\varphi'_1(\bar{X},\bar{x}) = \varphi_1(\bar{X},\bar{x})\wedge\psi_1(\bar{X},\bar{x}).
$$
This models that all the assignments that satisfy $\varphi_1$ also satisfy $\varphi'_1$, except for $(V^*,v^*)$. Then we define:
$$
\varphi'_2(\bar{X},\bar{x}) = \varphi_2(\bar{X},\bar{x})\wedge\psi_1(\bar{X},\bar{x})\wedge(\chi_1\to\psi_2(\bar{X},\bar{x})).
$$
This models that all the assignments that satisfy $\varphi_2$ also satisfy $\varphi'_2$, but if there was no assignment for $\varphi_1$, then we exclude $(V^*,v^*)$ for $\varphi_2$. Then we generalize this construction for each $\varphi_i$:
$$
\varphi_i'(\bar{X},\bar{x}) = \varphi_i(\bar{X},\bar{x})\wedge\psi_1(\bar{X},\bar{x})\wedge(\chi_1\to\psi_2(\bar{X},\bar{x}))\wedge((\chi_1\wedge\chi_2)\to\psi_3(\bar{X},\bar{x}))\wedge\cdots\wedge(
\bigwedge_{j = 1}^{j = i-1}\chi_j\to\psi_i(\bar{X},\bar{x})),
$$
By the construction of $\kappa(\alpha)$, one can easily check that  $\sem{\kappa(\alpha)}(\A) = \sem{\alpha}(\A)-1$ for each $\A$.
